### 2013-03-15 doc/sosp13/memory.tex
Modifications and notes in memory model/delegates section.

Under the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, Grappa caches are
\emph{not} kept coherent automatically. If that is desired, the programmer to
is responsible for pushing the updated block back to its home node. Given that
synchronization is explicit in Grappa, we don't expect this to be a
significant burden (programmers can push updates before a relevant
synchronization operation).

@@

Delegate operations are always executed at the home core of their
address, and while arbitrary memory operations can be delegated, we
restrict the use of delegate operations in three ways to make them more
useful for synchronization. First, we limit each task to one outstanding
delegate operation to avoid the possibility of reordering in the
network. Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the linear
address space so they can be satisfied with a single network request.
Finally, no context switches are allowed while the data is being
modified. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using actual atomic operations (and thus avoiding their typical high cost). 
Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
cache operations interact.

\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}

@@

\paragraph{Memory consistency model discussion} As mentioned earlier, all
synchronization operations are done via delegate operations. Since they all
execute in their home node, they are guaranteed to be serialized , with their
updates visible to all cores across the system in the same order. Also, tasks
are restricted to a single outstanding delegate operation at a time, and
therefore are not subject to potential request reordering. Consequently, all
synchronization operations execute in program order and are made visible in
the same order to all cores in the system. These properties are sufficient to
guarantee a memory model that offers sequential consistency for data-race-free
programs~\cite{AdveHill1990} (all accesses to shared data are separated by
synchronization). This is the memory model that underpins C/C++~\cite{N2480,N2800}.

Note, however, that if the application code uses explicit caching on shared
data, all updates must be sent back to the home node before the
synchronization operation that protects the data is performed.

>>>>

Under the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, Grappa caches are
\emph{not} kept coherent automatically. If that is desired, the programmer
is responsible for ensuring sufficient synchronization (via delegate
operations) occurs before acquiring or releasing cache objects.
% Given that synchronization is explicit in Grappa, we don't expect this to be a
% significant burden (programmers can push updates before a relevant
% synchronization operation).

@@

Delegate operations are always executed at the home core of their
address, and while arbitrary memory operations can be delegated, we
restrict the use of delegate operations in three ways to make them more
useful for synchronization.
First, the remote operation may not perform any operations that could cause a context switch; this ensures any modifications are atomic due to cooperative multithreading.
Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the linear
address space so they can be satisfied with a single network request.
Finally, the task issuing the delegate call blocks until the delegate operation
completes to ensure that synchronization has finished before continuing.
Using a ``futures''-like mechanism, tasks may issue delegates in parallel and block 
on the ``promises'' returned by asynchronous delegate invocations.
Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using actual atomic operations (and thus avoiding their typical high cost). 
Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
cache operations interact.

\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
\TODO{If we do want a section on GlobalCompletionEvent:
  \begin{itemize}
    \item many tasks and operations in the applications we've implemented do not need immediate consistency
    \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
    \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
    \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
    \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
    \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
    \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
    \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach
    \item interface to GlobalCompletionEvent:
      \begin{itemize}
        \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
        \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
        \item \texttt{gce.wait()}: block until all cores have no work
      \end{itemize}
    \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
  \end{itemize}
}

@@

\paragraph{Memory consistency model discussion} As mentioned earlier, all
synchronization operations are done via delegate operations. Since they all
execute in their home node, they are guaranteed to be serialized , with their
updates visible to all cores across the system in the same order. Also, tasks
are restricted to a single outstanding delegate operation at a time, and
therefore are not subject to potential request reordering \TODO{I think we can support multiple delegates in parallel from a task as long as we block on them before counting on them being complete. see my description above about the \emph{future/promise} delegates we have now. Not sure if there's a strong example of when this is useful for synchronization (though it's definitely useful for reads/writes), acquiring multiple locks doesn't work because we have to acquire locks in order to prevent deadlock.} Consequently, all
synchronization operations execute in program order and are made visible in
the same order to all cores in the system. These properties are sufficient to
guarantee a memory model that offers sequential consistency for data-race-free
programs~\cite{AdveHill1990} (all accesses to shared data are separated by
synchronization). This is the memory model that underpins C/C++~\cite{N2480,N2800}.

Note, however, that if the application code uses explicit caching on shared
data, all updates must be sent back to the home node before the
synchronization operation that protects the data is performed.\TODO{Not sure this warrants its own paragraph, seems like the explicit caches are just the same as reading and then writing using delegates.}
